---
title: "Data605 Assignment2"
author: 'jfor: Josh Forster'
date: "2023-02-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(pracma)
library(tidyverse)
```

(1) Show that \[\mathbf{A}^T \mathbf{A} \neq \mathbf{A} \mathbf{A}^T\] in general. (Proof and demonstration.)

1) IF \[m \neq n\] then the dot product will result in different row and column dimensions

\[\mathbf{A}  {m\times n} \times \mathbf{A}  {n\times m} = \mathbf{B} {m\times m} \]
\[
\mathbf{A}  {n\times m} \times \mathbf{A}  {m\times n} = \mathbf{B} {n\times n} \]

```{r m_equal_to_n}
a<- matrix(c(1,2,3,4,2,3,4,5,4,5,6,7),ncol=3,nrow=4)
print(a %*% t(a))
print(t(a) %*% a)

```

2) IF \[m = n\] 

given the transformation that occurs with matrix multiplication if \[1\leq i\leq n \neq 1\leq j\leq n\] for one non-diagnoal value the dot product will result in a different resulting matrix

```{r}
square <- matrix(c(1,2,3,4,5,6,7,8,9),ncol=3,nrow=3)
print(square %*% t(square))
print(t(square)%*%square)

```

(2) For a special type of square matrix A, we get \[\mathbf{A}^T \mathbf{A} = \mathbf{A} \mathbf{A}^T\] . Under what conditions
could this be true? (Hint: The Identity matrix I is an example of such a matrix).

The symmetric matrix is the only type of matrix where this conditions will hold true and   the identity matrix is a one example of this type of matrix. The matrix is equal to it's transpose \[\mathbf{A}^T = \mathbf{A} \] so it is as if this is squaring the matrix \[\mathbf{A}^2\]

```{rsymmetric example from text}
sym_mat <- matrix(c(2,3,-9,5,7,3,1,6,-2,-3,-9,6,0,-1,9,5,-2,-1,4,-8,7,-3,9,-8,-3),nrow=5,ncol=5)
print(sym_mat %*% t(sym_mat) == t(sym_mat) %*% sym_mat)
```

Matrix factorization is a very important problem. There are supercomputers built just to do matrix factorizations. Every second you are on an airplane, matrices are being factorized. Radars that track flights use a technique called Kalman filtering. At the heart of Kalman Filtering is a Matrix Factorization operation. Kalman Filters are solving linear systems of equations when they track your flight using radars.
Write an R function to factorize a square matrix A into LU or LDU, whichever you prefer.

```{r}
mat2 <- matrix(c(1,-2,3,4,8,4,-3,5,7),ncol=3,nrow=3,byrow=FALSE)
pracma::rref(mat2)
mat3 <- matrix(c(2,1,-6,4,-4,-9,-4,3,5),ncol=3,nrow=3,byrow=FALSE)
pracma::rref(mat3)
```

#### LU Decomposition

Each step depicts the row equivalent matrix as the row operation is applied to the original inputted matrix.

```{r lu_decomposition}

find_row_op <- function(vec,div,row_pos,col_pos,row_order){

    ops <- data.frame(cbind(-div/vec,row_order,rep.int(row_pos,length(vec)),rep.int(col_pos,length(vec))))
    colnames(ops)<-c('multiple','row_used','row_applied','col_applied')
    ops <- ops |> filter(row_used != row_pos) |> 
        mutate(first_rank=if_else(multiple==0,3,if_else(multiple %% 1==0,1,nchar(str_replace(as.character(multiple),'\\-?\\d.(\\d+)','\\1'))+1))) |>
        arrange(first_rank,abs(multiple)) |> 
        head(1)
    
    return(ops)
}

apply_row_op<- function(df,matr){
   return(matr[df$row_used,] * df$multiple + matr[df$row_applied,])
}

lu_decomp<- function(mata) {
    if (det(mata)==0) {
        print("This matrix cannot be inverted and cannot be transformed using LU decomposition")
    } else {
    counter = 1
    upper <- mata #[,]
    rownames(upper) <- seq(1,nrow(upper))
    upper_order <- cbind(upper,seq(1,nrow(upper)))
    lower <- diag(x=1,nrow=nrow(upper),ncol=ncol(upper))
    row_ops <- data.frame()
    for (j in 1:ncol(upper)){
    
        for (i in 1:nrow(upper)){
            if (i>j & upper[i,j] != 0) {
                if (nrow(row_ops)==0) {
                    cur_row_ops <- find_row_op(vec=upper[,j],div=upper[i,j],row_pos=i,col_pos=j,row_order=c(as.numeric(unlist(rownames(upper))))) 
                } else if (nrow(row_ops |> filter(row_applied == i))>0) {
                     sub_df <- row_ops |> filter(row_applied == i)
                     sub_rows = upper[upper[,max(sub_df$col_applied)]==0,]
                     sub_mat <- upper[c(rownames(sub_rows)),]

                     cur_row_ops <- find_row_op(vec=sub_mat[,j],div=upper[i,j],row_pos=i,col_pos=j,row_order=c(as.numeric(unlist(rownames(sub_mat)))))
                } else {
                    cur_row_ops <- find_row_op(vec=upper[,j],div=upper[i,j],row_pos=i,col_pos=j,row_order=c(as.numeric(unlist(rownames(upper)))))
                }
                cur_row_ops$order <- counter
                print(cur_row_ops)
                lower[cur_row_ops$row_applied,cur_row_ops$col_applied] <- -cur_row_ops$multiple
                row_ops <- rbind(row_ops,cur_row_ops)
                upper[cur_row_ops$row_applied,] <- apply_row_op(df=cur_row_ops,matr=upper)
                counter = counter + 1
                print(upper)
            }
        }
    }
    return(list(lower,upper)) 
    }
    }
```


#### 3x3 Matrix Testing

```{r function_testing}
mat2_orig <- matrix(c(1,-2,3,4,8,4,-3,5,7),ncol=3,nrow=3,byrow=FALSE)
mat2 <- matrix(c(2,1,-6,4,-4,-9,-4,3,5),ncol=3,nrow=3,byrow=FALSE)
test <- matrix(c(4,6,5,6),nrow=2,ncol=2)
t_results <- lu_decomp(mat2_orig)
print(t_results[[1]])

print(t_results[[2]])

print(t_results[[1]] %*% t_results[[2]])

print(t_results[[1]] %*% t_results[[2]] == mat2_orig)
```

#### 4x4 Matrix Testing

```{r}

mat_ex <- matrix(c(2,2,-3,6,2,1,-5,5,-2,6,-1,3,8,-3,2,5),nrow=4,ncol=4)
print(mat_ex)
lu <- lu_decomp(mat_ex)
print('Lower:')
print(lu[[1]]) 
print('Upper:') 
print(lu[[2]])

lu[[1]] %*% lu[[2]]
```

#### Additional Matrix Testing

```{r}
mat_ex2 <- matrix(c(4,2,-3,6,-4,1,-3,2,-2,6,-1,3,5,-3,2,5),nrow=4,ncol=4)
print(mat_ex2)
lu2 <- lu_decomp(mat_ex2)
print('Lower:')
print(lu2[[1]]) 
print('Upper:') 
print(lu2[[2]])

print(lu2[[1]] %*% lu2[[2]])

```

I believe some of the discrepancy in the end result when applying my LU decomposition \[ \mathbf{A} = \mathbf{L}\mathbf{U} \] is due to some rounding after applying row operations in the lower and upper matrices that causes differences after matrix multiplication.


